taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-10-03 paper/manuscript.Rmd
Add citations, render them in table.

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)

@@

printtable <- function(df){
  df %>% 
    kableExtra::kable("latex", booktabs=T) %>% 
    kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

- queries that return vectors: `get_ids()` and it's complement, `get_names()`,
- queries that filter the underlying taxonomic data frames: `by_name()`, `by_rank()`, `by_id()`, and `by_common()`,
- database functions  `td_create()`, `td_connect()` and `taxa_tbl()`,
- and helper utilities, such as `clean_names()`.

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette).

@@

## `by_` functions for access to underlying tables

Underlying data tables can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank. These functions allow us to ask general questions such as, how many bird species are there?

```{r}
by_rank("Aves", rank="class", provider = "col") %>%
  pull(taxonID) %>%
  n_distinct()
```

@@

```{r}
## example of sci name matching to more than one id.
by_name("Zosterops rendovae") %>% 
  select(1:5)

@@

`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names come with many of the same problems as scientific names but often more frequent, e.g. matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc. Common names are accessed via `by_common` which takes a vector of common names. The user can then resolve discrepencies.

@@

```{r cache = FALSE, echo = FALSE}
tibble(provider = c("Integrated Taxonomic Information System", 
                                 "National Center for Biological Information's Taxonomy database",
                                 "Catalogue of Life",
                                 "Global Biodiversity Information Facility Taxonomic Backbone",
                                 "FishBase",
                                 #"SeaLifeBase",
                                 "Wikidata", "Open Tree Taxonomy",
                                 "International Union for Conservation of Nature and Natural Resources"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
                                     "wd", "ott", "iucn"),
                    total_identifiers = map(abbreviation, 
                                            ~taxa_tbl(.x) %>% 
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
) %>% printtable()

>>>>

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)
library(knitr)

@@

printtable <- function(df){
  df %>%
    kableExtra::kable("latex", booktabs=T) %>%
    kableExtra::kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

- queries that return vectors: `get_ids()` and it's complement, `get_names()`,
- queries that filter the underlying taxonomic data frames: `filter_name()`, `filter_rank()`, `filter_id()`, and `filter_common()`,
- database functions  `td_create()`, `td_connect()` and `taxa_tbl()`,
- and helper utilities, such as `clean_names()`.

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette).

@@

## `filter_` functions for access to underlying tables

Underlying data tables can be accessed through the family of `filter_` functions, which filter by certain attributes such as scientific name, id, common name, and rank. These functions allow us to ask general questions such as, how many bird species are there?

```{r}
filter_rank("Aves", rank="class", provider = "col") %>%
  pull(taxonID) %>%
  n_distinct()
```

@@

```{r}
## example of sci name matching to more than one id.
filter_name("Zosterops rendovae") %>% 
  select(1:5)

@@

`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names come with many of the same problems as scientific names but often more frequent, e.g. matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc. Common names are accessed via `filter_common` which takes a vector of common names. The user can then resolve discrepencies.

@@

(ref:itis) Integrated Taxonomic Information System [@itis]

```{r echo = FALSE, results = "asis"}
tibble(provider = c("Integrated Taxonomic Information System, @itis",
                                 "National Center for Biological Information's Taxonomy database, @ncbi",
                                 "Catalogue of Life, @col",
                                 "Global Biodiversity Information Facility Taxonomic Backbone, @gbif",
                                 "FishBase, @fishbase",
                                 #"SeaLifeBase",
                                 "Wikidata, @jorrit", "Open Tree Taxonomy, @Rees2017",
                                 "International Union for Conservation of Nature and Natural Resources, @iucn"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
                                     "wd", "ott", "iucn"),
                    total_identifiers = map(abbreviation, 
                                            ~taxa_tbl(.x) %>% 
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
) %>% knitr::kable(format = "markdown")
